Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-29 survey/survey.tex
consistency up to 3.4. Write 3.5.

@@

Combinators of \ensuremath{\Pi } can be written in terms of the operators
described previously or via a graphical language
similar in spirit to those developed for Geometry of Interaction
\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
\cite{selinger-graphical}. Modulo some conventions and shorthand we
describe here, the wiring diagrams are equivalent to the operator
based (syntactic) description of programs.  \ensuremath{\Pi } combinators
expressed in this graphical language look like ``wiring
diagrams''. Values take the form of ``particles'' that flow along the
wires. Computation is expressed by the flow of particles.

@@

\noindent
\textit{Example.} As an example consider the wiring diagram
of the combinator \ensuremath{c} below:

\ensuremath{c : b \times (1+1) \leftrightarrow b + b}
\ensuremath{c = \swapt \odot \dist \odot (\identlt + \identlt)}
\begin{center}
\scalebox{1.5}{
%%subcode-line{pdfimage}[diagrams/example1-crop.pdf]
\includegraphics{inkscape/example1.pdf}
}
\end{center}

\jc{WAVEFRONT of consistency pass}
The textual representation of $\Pi$ is quite cumbersome: the
underlying principles however become quite intuitive when represented
in a more suggestive visual notation. There exist various graphical
notations for the monoidal categories that model $\Pi$ and related
languages including Penrose diagrams~\cite{selinger-graphical} and
Geometry of Interaction (GoI) machines and Proof
Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In this section, we
propose a novel graphical representation of $\Pi$ types and
combinators drawn from the resource usage embedded in the operational
semantics. As suggested by the operational semantics, the type $A * B$
can be viewed as denoting \emph{spatial composition}, where
\emph{both} $A$ and $B$ must happen in \emph{some} order or perhaps
simultaneously, and where $A + B$ denotes \emph{temporal composition},
where exactly \emph{one of} $A$ and $B$ must happen at a given time.
\jc{I think the temporal aspect is stretching things too far. Could we
call it ``choice composition''? (Or something better). To me spatial
composition somehow seems like sequential composition (i.e. composition
in time), which is definitely not what we want. I think the informal
description (both although perhaps concurrently, and exactly one of)
is better. In section 3 we called them sequential, choice and
parallel. We should be consistent.}

@@

\jc{
\texttt{c2equiv} combinator in the same file, which interprets
every combinator as a type equivalence, which can then just be
run. \texttt{lemma0} and \texttt{lemma1} in the same file show
that this is indeed the case. The advantage is that this lifts
to level 2 quite cleanly.}

>>>>

%%%%%%%%%
\subsection{Operational Semantics}
\label{sec:opsem}

@@

Combinators of \ensuremath{\Pi } can be written in terms of the operators
described previously or via a graphical language
similar in spirit to those developed for Geometry of Interaction
\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
\cite{BLUTE1996229,selinger-graphical}. Modulo some conventions and shorthand we
describe here, the wiring diagrams are equivalent to the operator
based (syntactic) description of programs.  \ensuremath{\Pi } combinators
expressed in this graphical language look like ``wiring
diagrams''. Values take the form of ``particles'' that flow along the
wires. Computation is expressed by the flow of particles.

@@

\noindent
\textit{Example.} As an example consider the wiring diagram
of the combinator \ensuremath{c} below:
\begin{flalign*}
& \begin{aligned}
c & : & b \times (1+1) \leftrightarrow b + b \\
c & = & \swapt \odot \dist \odot (\identlt + \identlt) \\
\end{aligned} &&
\end{flalign*}

\begin{center}
\scalebox{1.5}{
%%subcode-line{pdfimage}[diagrams/example1-crop.pdf]
\includegraphics{inkscape/example1.pdf}
}
\end{center}

@@

Figure~\ref{type-isos} introduces our desired denotational
semantices, and section~\ref{sec:opsem} a direct definition
of an operational semantics. One obvious question arises:
do these correspond?

We can certainly associate to each $\Pi$ combinator an
equivalence between the denotation of each type%
\footnote{This is extracted from the Agda formalization
of this work, which has been reported on in~\cite{Carette2016}.}:
\begin{code}\hspace*{-4mm}
\>\AgdaFunction{c2equiv} \AgdaSymbol{:} \AgdaSymbol{\{}\AgdaBound{t₁}
\AgdaBound{t₂} \AgdaSymbol{:} \AgdaDatatype{U}\AgdaSymbol{\}} \AgdaSymbol{→}
\AgdaSymbol{(}\AgdaBound{c} \AgdaSymbol{:} \AgdaBound{t₁} \AgdaDatatype{⟷}
\AgdaBound{t₂}\AgdaSymbol{)} \AgdaSymbol{→} \AgdaFunction{⟦} \AgdaBound{t₁}
\AgdaFunction{⟧} \AgdaFunction{≃} \AgdaFunction{⟦} \AgdaBound{t₂}
\AgdaFunction{⟧}%
\end{code}

\noindent And as such an equivalence contains a function as
its first component, we can compare if our operational
semantics and denotational semantics match.  And they do:
\begin{code}\hspace*{-4mm}
\>\AgdaFunction{lemma0} \AgdaSymbol{:} \<[10]%
\>[10]\AgdaSymbol{\{}\AgdaBound{t₁} \AgdaBound{t₂} \AgdaSymbol{:}
\AgdaDatatype{U}\AgdaSymbol{\}} \AgdaSymbol{→} \AgdaSymbol{(}\AgdaBound{c}
\AgdaSymbol{:} \AgdaBound{t₁} \AgdaDatatype{⟷} \AgdaBound{t₂}\AgdaSymbol{)}
\AgdaSymbol{→} \AgdaSymbol{(}\AgdaBound{v} \AgdaSymbol{:} \AgdaFunction{⟦}
\AgdaBound{t₁} \AgdaFunction{⟧}\AgdaSymbol{)} \AgdaSymbol{→}
\AgdaFunction{eval} \AgdaBound{c} \AgdaBound{v} \AgdaDatatype{≡}
\AgdaField{proj₁} \AgdaSymbol{(}\AgdaFunction{c2equiv}
\AgdaBound{c}\AgdaSymbol{)} \AgdaBound{v}\<%
\end{code}

\noindent We can similarly hand-write a backwards evaluator,
prove that it is indeed a proper backwards evaluator, and
finally show that it agrees with the reverse equivalence.
